The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-07-20 VegOzone.tex
Revision of introduction.

Tropospheric and in particular ambient air ozone is a highly toxic pollutant to all ecosystems. In continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is cause to early senescence as well as programmed cell death in plants {\bf TODO citations}. On average, a global loss of yield in four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Former can proof a vital threat to food security in fast developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone related to air pollution has increased in the past years. At the same time, a successful implementation of air quality regulations reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

@@

In the presents of nitrogen oxides (\chem{NO_x}), tropospheric ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs). In the absence of local pollution sources, episodes of high ozone concentrations are often attributed to long-range transport of precursors. In northern Fennoscandia episodes in 2003 and 2006 have been traced back to emissions related to forest fires in southern Europe and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. But also living vegetation emits hydrocarbons in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). Those emissions are associated with, e.g., thermal stress, defense against herbivores, or attraction of pollinators \citep{TPS:Penuelas2003}.

@@

The opening of stomata, described through their conductance to gases $g_\mathrm{sto}$, is governed by different environmental factors and is described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$. Stomata are typically considered closed below a light intensity of $50\,\unit{W\,m^{-2}}$.

To measure the risk of damage inflicted to vegetation through ozone, many different metrics have been defined. Two main types can be differentiated, concentration-based (e.g., $\mathrm{AOT_{40}}$) and flux-based (e.g.,$\mathrm{POD_{0.1}}$) metrics. Due to their simplicity, $\mathrm{AOT_{x}}$ are still widely used, despite the actual integrated flux of ozone entering through the stomata is more relevant for an estimate of the actual toxic dose. 
The Accumulated Ozone Exposure over a threshold concentration of $x\,\unit{ppb}$ is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. The integration time varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.
The Phototoxic Ozone Dose (POD) over a threshold y is defined as:
\begin{equation}
  \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$. The stomatal conductance is either measured directly or computed in accordance to Eq.~(\ref{eq:stomatal}).\\

@@

Heat and drought stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high pressure system extended even over a period of five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors (\chem{CO}, \chem{VOCs}).

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}.
Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.

@@

In terms of AOT40, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019}. {\bf TODO: come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia. Later supplement figure \ref{fig:fennoscandic_aot40}}

>>>>

Tropospheric and in particular ambient air ozone is a highly toxic pollutant to all ecosystems. In continuously high concentration, ozone negatively affects life expectancy of humans \citep{WHO2008} and is cause to early senescence as well as programmed cell death in plants \citep{PCE:Kangasjarvi2005}. On average, a global loss of yield in four major crops (wheat, rice, maize, and soybean) of $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} and a loss in primary production in forestry of $7\,\unit{\%}$ \citep{GCB:Wittig2009,EP:Matyssek2012} has been attributed to ozone. Former can proof a vital threat to food security in fast developing countries, e.g., in East and South-East Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018}, where ozone related to air pollution has increased in the past years. At the same time, a successful implementation of air quality regulations reduced the number of episodes of peak concentrations especially in summertime over Europe and North America \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}.

@@

In the presents of nitrogen oxides (\chem{NO_x}), tropospheric ozone is formed in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) and hydrocarbons known as volatile organic substances (VOCs). In the absence of local pollution sources, episodes of high ozone concentrations are often attributed to long-range transport of precursors. In northern Fennoscandia, episodes in 2003 and 2006 have been traced back to emissions related to forest fires in southern and eastern Europe, respectively \citep{AE:Lindskog2007,EP:Karlsson2013}. Hydrocarbons are also emitted by living vegetation in form of terpenes and monoterpenes, so called biogenic volatile substances (BVOCs). Those emissions are associated with, e.g., thermal stress, defense against herbivores, or attraction of pollinators \citep{TPS:Penuelas2003}.

@@

The opening of stomata, typically quantified through a conductance to water vapor $g_\mathrm{sto}$, is governed by different environmental factors and is described by a multiplicative ansatz \citep{PTRS:Jarvis1976,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within a range $0-1$. They are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type (PFT) $k$. Stomata are typically considered closed below a light intensity of $50\,\unit{W\,m^{-2}}$.

To measure the risk of damage inflicted to vegetation through ozone, many different metrics have been defined. Two main types can be differentiated, concentration-based (e.g., $\mathrm{AOT_{40}}$) and flux-based (e.g.,$\mathrm{POD_{0.1}}$) metrics. Due to their simplicity, $\mathrm{AOT_{x}}$ are still widely used, despite the actual integrated flux of ozone entering through the stomata is more relevant for an estimate of the actual toxic dose \citep{ICP:MappingManual2017}. 
The Accumulated Ozone Exposure over a threshold concentration of $\mathrm{x}$ in units of $\unit{ppb}$ is defined as:
\begin{equation}
  \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \label{eq:aot}
\end{equation}
with the ambient ozone concentration $\chem{[O_3]}$. The integration time varies, but is usually defined from $8\,\unit{am}-8\,\unit{pm}$.
The Phototoxic Ozone Dose (POD) over a threshold y is defined as:
\begin{equation}
  \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
  \label{eq:pod}
\end{equation}
with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$. The stomatal conductance is either measured directly or computed in accordance to Eq.~(\ref{eq:stomatal}).\\

@@

Heat and drought stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high pressure system extended even over a period of five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and \chem{VOCs} ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.

A total of 2079 forest fires of different size were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer) and an estimated total of $25000\,\unit{hectare}$ burned down in the course of the most extensive fires in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) in younger history \citep{SOU2019}.

@@

In terms of $\mathrm{AOT_{40}}$, the threshold for protection of vegetation of $3000\,\unit{ppb}$ (integrated over the growing season) given by the EU directive has been well exceeded at many ozone monitoring sites in Southern Norway in the 2018 growing season \citep{NILU2019}. {\bf TODO: come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia. Later supplement figure \ref{fig:fennoscandic_aot40}}
